---
title: "Bayesian Analysis of Syntopy"
author: "Matěj Tvarůžka"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# Introduction

Following the framework of Remeš & Harmáčková (2023), this assignment simulates and analyses local co-occurrence (syntopy) among sister species of birds. In evolutionary ecology, syntopy is the final stage of the speciation cycle, during which two sister species meet at the same sites. I model the co-occurrence probability at local sites as a function of ecological and spatial predictors. This idea is connected to the topic of my master's thesis, where I model syntopy at two spatial scales (routes and stops) using BBS (Breeding Bird Survey) data for passerines and range maps from eBird Status & Trends.

To measure syntopy accurately, we need to consider only sites within the range overlap for the analysis. We can imagine the range overlap being the jar, from which we randomly draw balls (sites). There are four possible outcomes of that draw: 1) both species are present, 2) species A is present, 3) species B is present, or 4) both species are missing at the site. To measure the degree of species co-occurrence, we use the log-odds ratio because it centers the odds ratio around zero, making the scale linear.

If we want to model this problem using Bayesian methods, we can follow the approach in Newbury (2024), but we will not include random effects in the model and will use only a simple Bayesian linear regression. The most accurate probability distribution for our problem is **Fisher's non-central hypergeometric distribution**, where *N* denotes the population size (number of sites within the sympatric zone), *K* is the number of co-occurrence states (*K* = 4), *n* is the number of draws, and *k* is the number of co-occurrences. This distribution is accurate because the sampling is carried out without replacement (unlike the binomial distribution), which reflects the true nature of the problem, given the limited number of pairs.

::: callout-note
## Goals

1.  Generate dream data for 100 pairs using Fisher's hypergeometric distribution for both scales (routes and stops).
2.  Visualize and explore generated data.
3.  Fit models for both scales using `ulam` function and compare the estimates with true values.
:::

# Data Simulation

To satisfy the assignment's complexity requirements, I define **5 true parameters** that dictate the data-generating process:

- $\beta_{0} = 0.5$ (Baseline log-odds of co-occurrence; slight facilitation)

- $\beta_{res.div} = 0.8$ (Resource use divergence: greater divergence increases syntopy)

- $\beta_{hab.pref} = -0.6$ (Difference in micro-habitat and environmental preference decreases syntopy)

- $\beta_{prod} = 0.3$ (Environmental productivity (e.g., NDVI): richer environments support co-occurrence)

- $\beta_{sympatry} = 1.2$ (Spatial overlap: regional overlap strongly increases syntopy)

## Fishers' Hypergeometric Distribution

### Mathematical Formulation

For each site $i$ out of $N$, the co-occurrence affinity ($α$) is the log-odds ratio of given parameters:

$$\text{log}(ω_i) = \beta_{0} + \beta_{res.div} \cdot X_{1,i} + \beta_{hab.pref} \cdot X_{2,i} + \beta_{prod} \cdot X_{3,i} + \beta_{sympatry} \cdot X_{4,i}$$

The observed co-occurrence $k_i$ is modeled with **Fisher's Hypergeometric distribution** representing local survey plots within that site:

$$k_i \sim \text{FisherHypergeometric}(m_{1i},m_{2i}, N_i, ω_i)$$

where:

- $k_i$ = number of sites where both species co-occur

- $m_{1i}$ = total number of sites where species A is present for pair $i$

- $m_{2i}$ = total number of sites where species B is present for pair $i$

- $N_i$ = total number of sites sampled within the sympatric range of pair $i$

- $ω_i$ = the conditional co-occurrence odds ratio for pair $i$

## Data Generation

I want to simulate data resembling the study design of BBS (\~40km long routes, each having 50 stops). Therefore I need to create two data sets:

1\) route scale (looking at syntopy across transects within the range overlap)

2\) stop scale (looking at syntopy within transects)

For the Fisher's hypergeometric distribution we can use the library `BiasedUrn`, for the modeling we will use `rethinking` and `rstan`.

```{r}
#| include: false
library(BiasedUrn)
library(rethinking)
library(rstan)
library(tidyverse)
set.seed(1)
```

### Route Scale

For the number of sites I choose a negative binomial distribution, because it is an integer and I know that there is a long tail on the right site of the distribution.

```{r}
#| label: Route Scale Data Simulation
num_pairs <- 100
route_occupancy <- function(num_pairs){
  # simulates the occupancy of species across N routes
  
  # N of routes
  N <-  rnbinom(n = num_pairs, mu = num_pairs, size = 2)
  # Probability distribution for species occupancy
  prob_m1 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  prob_m2 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  # Species occupancy
  m1 <- rbinom(num_pairs, size = N, prob = prob_m1)
  m2 <- rbinom(num_pairs, size = N, prob = prob_m2)
  list(N=N, m1=m1, m2=m2)
}
route_oc <- route_occupancy(num_pairs = num_pairs)

hist(route_oc$N, breaks = 12, main="N Sites Distribution", xlab="Number of sites for sister pair", col="purple")
```

Some species pairs have very large number of sampled sites, while some of them have just a few sampled sites within the sympatric range. We can check the parameters of the distribution:

```{r}
summary(route_oc$N)
```

We can also check the occupancy distribution of the first species set in the pairs:

```{r}
hist(route_oc$m1, breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

This plot nicely resembles species abundance distribution, so if we take a logarithm of it it should be close to normal distribution.

```{r}
hist(log(route_oc$m1), breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

It is exactly like that!

Let's continue with the generation of parameter values. First I will generate three already standardized variables using normal distribution.

```{r}
X_res_div   <- rnorm(num_pairs, mean = 0, sd = 1) # X1
X_hab_pref  <- rnorm(num_pairs, mean = 0, sd = 1) # X2
X_prod      <- rnorm(num_pairs, mean = 0, sd = 1) # X3
hist(X_res_div, breaks = 12, main="Resource Use Divergence", xlab="Degree of divergence", col="orange")
```

For the degree of sympatry (% of range overlap) I need to generate specific distribution as x \> 5 (minimum degree for sympatry is typically 5 %) and x \<= 100. As I want the values to have decimals and I need a strict upper limit I have to use beta distribution.

```{r}
raw_beta <- rbeta(num_pairs, shape1 = 2, shape2 =3)
sym_min <- 5 # sympatry threshold is 5 %
symp_max <- 100 # full overlap
sympatry <- sym_min + (raw_beta - min(raw_beta)) * (symp_max - sym_min) / (max(raw_beta) - min(raw_beta))
hist(sympatry, breaks = 12, main="Degree of Sympatry", xlab="Range overlap (%)", col="darkseagreen3")
```

We can see that most of the pairs have a high degree of sympatry and few pairs have very low values. Now we have to standardize it prior to modeling.

```{r}
#| label: Parameter X4
X_sympatry <- (sympatry - mean(sympatry)) / sd(sympatry) # X3
hist(X_sympatry, breaks = 12, main="zSympatry (Standardized)", xlab="Z-score", col="steelblue")
```

Now we set the "true" values for the parameters in our made-up universe.

```{r}
#| label: True Parameters
beta_0        <- 0.5
beta_res_div  <- 0.8
beta_hab_pref <- -0.6
beta_sympatry <- 1.2
beta_prod     <- 0.3
```

Next step is the calculation of linear combination of parameters and we exponentiate the log-odds to get simple odds ratio.

```{r}
#| label: Odds Ratio
log_omega <- beta_0 + 
             beta_res_div  * X_res_div + 
             beta_hab_pref * X_hab_pref + 
             beta_sympatry * X_sympatry + 
             beta_prod     * X_prod
omega <- exp(log_omega)
```

With this settled up, we calculate the linear predictor for each pair using `BiasedUrn` library.

```{r}
#| label: Simulation of co-occurrence
m1 <- route_oc$m1
m2 <- route_oc$m2
k <- integer(num_pairs)
for(i in 1:num_pairs) {
  set.seed(1)
  # Simple safety boundary: if a species isn't present at all, co-occurrence is 0
  if(m1[i] == 0 || m2[i] == 0) {
    k[i] <- 0
  } else {
    # rFNCHypergeo(n_samples, m1, m2, total_draws, odds)
    k[i] <- rFNCHypergeo(1, route_oc$m1[i], route_oc$N[i] - route_oc$m1[i], route_oc$m2[i], omega[i])
  }
}
```

Let's pack it up into data.frame and take a look at the data.

```{r}
#| label: Route Scale Data Final
route_data <- data.frame(
  N = route_oc$N, m1 = m1, m2 = m2, k = k,
  res_div = X_res_div, hab_pref = X_hab_pref,
  prod = X_prod, 
  sympatry = X_sympatry
)
head(route_data)
```

The data.frame looks good and we should now explore the data before proceeding into next stage.

```{r}
#| label: Route Scale Data Exploration & Wrangling
nrow(route_data[route_data$k == 0,])
# There is 18 pairs with k=0
route_data[abs(route_data$m1 - route_data$m2) >100,]
# There are 4 pairs with a very high difference in occupancy between species
```

Higher abundance among sister pairs (e.g. 80 occupied routes from 100, together at 64) can lead to the situation when both species in the pair tend to co-occur more likely simply because they are present everywhere - therefore their co-occurrence affinity will be incomparable with the species which co-occur despite being rare in general (e.g. 5 routes from 100, together at 4).

This problem of prevalence can happen if we measure co-occurrence with widely used indices of similarity (e.g., Jaccard, Simpson). Bayesian inference elegantly solves this problem as it proportionately weights the relative abundances of species.

```{r}
# Add unique IDs, true odds ratio and true log-odds ratio
route_data$pair_id <- 1:nrow(route_data)
route_data$OR <- omega
route_data$log_OR <- log_omega
route_data[route_data$m1 == 0 & route_data$m2 == 0,]
```

There are 2 routes with zero occupancy, so called double zeros. Double zeros could mean two things. Either both species are truly absent at the route (true negatives) or they were not detected during the survey (false negatives). Given that route is within their range zone and there were 50 points where at least one of the species could have been detected, it should be more likely that they were simply not there. However, because these data are generated, we have pair's true odds ratio. If we would calculate odds ratio only from occupancy data, we would end up with infinity value.

### Stop Scale

Next step is to simulate point scale data based on the data for routes, which we already have. I choose to filter only pairs which have **k \>= 1**, because there is no reason to calculate syntopy on transect where is 0 possibility of co-occurrence.

```{r}
# Select only rows where k >= 1
valid_route_data <- route_data[route_data$k >= 1, ]

# Initialize empty list and counter for fast execution
datalist <- list()
counter <- 1

for (row_idx in 1:nrow(valid_route_data)) {
  
  # Extract pair metadata and parameters
  current_pair  <- valid_route_data$pair_id[row_idx]
  num_routes <- valid_route_data$k[row_idx]
  local_omega   <- valid_route_data$OR[row_idx] # using same omega as for the routes
  
  # Calculate the realized regional occupancy rates for this specific pair
  rate_A <- valid_route_data$m1[row_idx] / valid_route_data$N[row_idx]
  rate_B <- valid_route_data$m2[row_idx] / valid_route_data$N[row_idx]
  
  # Loop through each of those specific co-occurring transects
  for (t in 1:num_routes) {
    S_total <- 50
    
    # Generate local point occupancy probabilities using the pair's regional rates
    local_prob_A <- rbeta(1, shape1 = rate_A * 5, shape2 = (1 - rate_A) * 5)
    local_prob_B <- rbeta(1, shape1 = rate_B * 5, shape2 = (1 - rate_B) * 5)
    
    # Boundary protection to prevent 0 or 1 probabilities
    local_prob_A <- max(0.01, min(0.99, local_prob_A))
    local_prob_B <- max(0.01, min(0.99, local_prob_B))
    
    # Draw stop-scale occupancy counts bounded by 50 points
    s_m1 <- rbinom(1, size = S_total, prob = local_prob_A)
    s_m2 <- rbinom(1, size = S_total, prob = local_prob_B)
    
    # Safety floor: since they co-occurred on this transect, they must occupy >= 1 point
    s_m1 <- max(1, s_m1)
    s_m2 <- max(1, s_m2)
    
    # Draw realized stop co-occurrences using Fisher's Noncentral Hypergeometric
    s_k <- rFNCHypergeo(1, s_m1, S_total - s_m1, s_m2, local_omega)
    
    # Wrap it up in the list
    datalist[[counter]] <- data.frame(
      route_id = paste0("Pair_",current_pair, "_Route_", t),
      pair_id = current_pair,
      S_total = S_total,
      m1_stops = s_m1,
      m2_stops = s_m2,
      k_stops = s_k
    )
    counter <- counter + 1
  }
}

data <- do.call(rbind, datalist)
head(data)
```

Finally, we join our `data` with the generated parameters for each pair we already have in `route_data`.

```{r}
# Start with the simulated data (which naturally excludes k = 0 pairs)
stop_data <- data |>
  select(pair_id, route_id, S_total, m1_stops, m2_stops, k_stops) |> 
  left_join(
    # Pull ONLY the environmental covariates from route_data
    route_data |> select(pair_id, res_div, hab_pref, prod, sympatry), 
    by = "pair_id"
  )

# Verify the clean data frame
head(stop_data)
```

# Models

We have prepared our data and now we can run some models using the `ulam()` function from `rethinking` package. There are several steps we need to take to get the output:

1.  Wrap up data into list, so `ulam` can read it.

2.  Import the separate file for Fishers's non-central hypergeometric likelihood function (box 1).

3.  Set up the model formula in `ulam`.

4.  Transform the code into stan syntax.

5.  Run the model using `stan()` function from `rstan` package.

::: {.callout-tip collapse="false"}
## Box 1

The mathematical engine of this analysis is the custom **Fisher's Non-Central Hypergeometric Log Probability Mass Function (`fisher_hyper_lpmf`)**. Unlike a standard hypergeometric distribution, which assumes species sort completely at random, this distribution accounts for ecological attraction or repulsion via an odds ratio parameter.

### 1. Constraints and Support Bounds

The function first calculates the absolute physical boundaries of co-occurrence ($k$) allowed by the individual species' occupancy counts ($m_1, m_2$) within the total sampled sites ($N$): $$k_{\min} = \max(0, m_1 + m_2 - N)$$ $$k_{\max} = \min(m_1, m_2)$$

If an observed or sampled co-occurrence count falls outside this range ($k < k_{\min}$ or $k > k_{\max}$), the function returns $-\infty$, assigning it a probability of 0 and rendering that parameter space impossible for the MCMC sampler.

### 2. The Log-Numerator (The Observed State)

For the actual observed co-occurrences ($k$) under a estimated conditional odds ratio ($\mu$), the unnormalized log-probability is calculated using log-combinatorics (`lchoose`): $$\log(\text{Numerator}) = \ln \binom{m_1}{k} + \ln \binom{N - m_1}{m_2 - k} + k \cdot \ln(\mu)$$

### 3. The Partition Function (Normalization Constant)

Because the probabilities of all possible outcomes must sum to exactly 1, the function loops through every valid state from $k_{\min}$ to $k_{\max}$ to calculate the total probability space vector ($\mathbf{T}$). It then normalizes the final probability by subtracting the total sum of this vector from our observed state using a numerically stable routine: $$\text{LPMF} = \log(\text{Numerator}) - \text{log\_sum\_exp}(\mathbf{T})$$

In a log-space framework, subtraction mathematically equates to division, successfully producing a clean, fully normalized probability density score for the sampler.
:::

## Route Scale Model

```{r}
dat_list <- list(
  k = route_data$k,
  m1 = route_data$m1,
  m2 = route_data$m2,
  N = route_data$N,
  res_div = route_data$res_div,
  hab_pref = route_data$hab_pref,
  prod = route_data$prod,
  sympatry = route_data$sympatry
)
# Read separate stan file and wrap it in the required Stan syntax
likelihood_pmf <- paste(
  "functions {",
  paste(readLines("fisher_nchyper_lpmf.stan"), collapse = "\n"),
  "}",
  sep = "\n"
)

# Model settings
model <- ulam(
  alist(
    # Likelihood
    k ~ custom( fisher_hyper_lpmf(k[i] | exp(log_omega[i]), m1[i], m2[i], N[i]) ),
    
    # Linear Predictor
    log_omega <- b0 + b_res*res_div + b_hab*hab_pref + b_prod*prod + b_sym*sympatry,
    
    # Priors
    b0 ~ dnorm(0, 10),
    b_res ~ dnorm(0, 10),
    b_hab ~ dnorm(0, 10),
    b_sym ~ dnorm(0, 10),
    b_prod ~ dnorm(0, 10)
  ), 
  data = dat_list, 
  sample = FALSE # Generate text for stan, not sampling yet
)

# Extract the autogenerated text
stan_code <- stancode(model)

# Paste the file-loaded functions block to the top of the text string
full_stan_code <- paste0(likelihood_pmf, "\n", stan_code)

# Run the compiled model using rstan
out_routes <- stan(
  model_code = full_stan_code,
  data = dat_list,
  chains = 4, 
  cores = 4,
  iter = 4000,
  seed = 1,
  refresh = 0 # <--- Turns off the iteration progress tracker
)
```

First we can check the convergence of chains.

```{r}
traceplot(out_routes, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod"))
```

All chains have converged nicely!

```{r}
precis(out_routes)
```

Our model estimates for most parameters are very good, however the intercept (baseline log-odds ratio) is a bit underestimated despite the credible interval includes the true value (0.5).

## Stop Scale Model

```{r}
# Map character route IDs to sequential integers for Stan indexing
stop_data$route_index <- as.integer(as.factor(stop_data$route_id))

dat_list_stops <- list(
  k_stops     = stop_data$k_stops,
  m1_stops    = stop_data$m1_stops,
  m2_stops    = stop_data$m2_stops,
  S_total     = stop_data$S_total,
  res_div     = stop_data$res_div,
  hab_pref    = stop_data$hab_pref,
  prod        = stop_data$prod,
  sympatry    = stop_data$sympatry,
  route_index = stop_data$route_index
)

# Model Settings
model_stops <- ulam(
  alist(
    # Likelihood using stop-level data arrays
    k_stops ~ custom( fisher_hyper_lpmf(k_stops[i] | exp(log_omega[i]), m1_stops[i], m2_stops[i], S_total[i]) ),
    
    # Linear Predictor: Pair baseline + Route random effect deviation
log_omega <- b0 + b_res*res_div + b_hab*hab_pref + b_prod*prod + b_sym*sympatry + (z_route[route_index] * sigma_route),

    # Fixed Effect Priors
    b0 ~ dnorm(0, 10),
    b_res ~ dnorm(0, 10),
    b_hab ~ dnorm(0, 10),
    b_sym ~ dnorm(0, 10),
    b_prod ~ dnorm(0, 10),
    
    # Adaptive Priors
    # Non-Centered Random Effect Layer
    # z_route is a completely standard, clean vector array
    z_route[route_index] ~ dnorm(0, 1),
    sigma_route ~ dexp(1)
    # exponential distribution forces sigma_route
    # to be strictly positive (since a standard deviation cannot be negative)
  ), 
  data = dat_list_stops, 
  sample = FALSE # Generate text for stan, not sampling yet
)

# Extract the autogenerated text
stan_code_stops <- stancode(model_stops)

# Paste the file-loaded functions block to the top of the text string
full_stan_code_stops <- paste0(likelihood_pmf, "\n", stan_code_stops)

# Run the compiled model using rstan
out_stops <- stan(
  model_code = full_stan_code_stops,
  data = dat_list_stops,
  chains = 4, 
  cores = 4,
  iter = 4000,
  seed = 1,
  refresh = 0 # <--- Turns off the iteration progress tracker
)
```

There were several warnings at the and of the sampling, that some chains did not converged in an optimal way so we first check the traceplot for the main parameters.

```{r}
traceplot(out_stops, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod", "sigma_route"))
```

Now we can take a look at the parameter estimates.

```{r}
precis(out_stops, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod", "sigma_route"))
```

```{r}
# Extract posterior sample draws
post_stops <- extract.samples(out_stops)
n_samples  <- length(post_stops$b0)
n_rows     <- nrow(stop_data)

# Create an empty matrix to hold log-odds (rows = post-samples, cols = stop_data rows)
log_omega_routes <- matrix(NA, nrow = n_samples, ncol = n_rows)

# Reconstruct log_omega per transect by pulling pair fixed effects and adding local random noise
for(i in 1:n_rows) {
  pair_base <- post_stops$b0 + 
               post_stops$b_res * stop_data$res_div[i] + 
               post_stops$b_hab * stop_data$hab_pref[i] + 
               post_stops$b_sym * stop_data$sympatry[i] + 
               post_stops$b_prod * stop_data$prod[i]
  
  # Manually reconstruct the route deviation from the non-centered samples
  current_route_idx <- stop_data$route_index[i]
  reconstructed_a_route <- post_stops$z_route[, current_route_idx] * post_stops$sigma_route
  
  # Total local log-odds ratio
  log_omega_routes[, i] <- pair_base + reconstructed_a_route
}

# 1. Compute and assign the estimated log-odds ratio back to each individual transect row
stop_data$estimated_local_log_OR <- colMeans(log_omega_routes)

# 2. Average the estimated log-odds values for each distinct pair_id
pair_level_averages <- aggregate(estimated_local_log_OR ~ pair_id, data = stop_data, FUN = mean)
print(head(pair_level_averages))
```

We can compare it with the true values of local logOR.

```{r}
stop_model_evaluation <- pair_level_averages %>% 
  left_join(route_data %>% select(pair_id, true_log_OR = log_OR), by = "pair_id")
print(head(stop_model_evaluation))
```

# Smaller number of sister pairs (n=30)

```{r}

```

```{r}

```

# Conclusion

# Reference

Remeš, V. and Harmáčková, L. (2023), Resource use divergence facilitates the evolution of secondary syntopy in a continental radiation of songbirds (Meliphagoidea): insights from unbiased co-occurrence analyses. Ecography, 2023: e06268. https://doi.org/10.1111/ecog.06268

Newbury, A. (2024). Bayesian estimation of co-occurrence affinity with dyadic regression. bioRxiv. https://doi.org/10.1101/2024.01.16.575941

Haldane, J. B. S. (1956). *The estimation and significance of the logarithm of a ratio of frequencies*. Annals of Human Genetics, 20(4), 309-311.

Anscombe, F. J. (1956). *On estimating binomially distributed populations*. Biometrika, 43(3/4), 461-464. :::
