---
title: "Bayesian Analysis of Syntopy"
author: "Matěj Tvarůžka"
date: "`r Sys.Date()`"
format: html
editor: visual
---

ADD comments to your code!!! Create some functions to make it easy for changing values! Can I use matplot() for comparing how the co-occurrence depends on N sites? use dim()

use system.time()

# Introduction

Following the framework of Remeš & Harmáčková (2023), this assignment simulates and analyses local co-occurrence (syntopy) among sister species of birds. In evolutionary ecology, syntopy is the final stage of the speciation cycle, during which two sister species meet at the same sites. I model the co-occurrence probability at local sites as a function of ecological and spatial predictors. This idea is connected to the topic of my master's thesis, where I model syntopy at two spatial scales using BBS (Breeding Bird Survey) data for passerines and range maps from eBird Status & Trends.

To measure syntopy accurately, we need to consider only sites within the range overlap for the analysis. We can imagine the range overlap being the jar, from which we randomly draw balls (sites). There are four possible outcomes of that draw: 1) both species are present, 2) species A is present, 3) species B is present, or 4) both species are missing at the site. To measure the degree of species co-occurrence, we use the log-odds ratio because it centers the odds ratio around zero, making the scale linear.

If we want to model this problem using Bayesian methods, we can follow the approach in Newbury (2024), but we will not include random effects in the model and will use only a simple Bayesian linear regression. The most accurate probability distribution for our problem is **Fisher's non-central hypergeometric distribution**, where *N* denotes the population size (number of sites within the sympatric zone), *K* is the number of co-occurrence states (*K* = 4), *n* is the number of draws, and *k* is the number of co-occurrences. This distribution is accurate because the sampling is carried out without replacement (unlike the binomial distribution), which reflects the true nature of the problem, given the limited number of pairs.

::: callout-note
## Goals

1.  Generate dream data for 100 pairs using Fisher's hypergeometric distribution for both scales.
2.  Visualize generated data.
3.  Fit model using `ulam` function and compare estimates using different N for both scales.
:::

# Data Simulation

To satisfy the assignment's complexity requirements, I define **5 true parameters** that dictate the data-generating process:

- $\beta_{0} = 0.5$ (Baseline log-odds of co-occurrence; slight facilitation)

- $\beta_{res.div} = 0.8$ (Resource use divergence: greater divergence increases syntopy)

- $\beta_{hab.pref} = -0.6$ (Difference in micro-habitat and environmental preference decreases syntopy)

- $\beta_{prod} = 0.3$ (Environmental productivity (e.g., NDVI): richer environments support co-occurrence)

- $\beta_{sympatry} = 1.2$ (Spatial overlap: regional overlap strongly increases syntopy)

## Fishers' Hypergeometric Distribution

### Mathematical Formulation

For each site $i$ out of $N$, the co-occurrence affinity ($α$) is the log-odds ratio of given parameters:

$$\text{log}(ω_i) = \beta_{0} + \beta_{res.div} \cdot X_{1,i} + \beta_{hab.pref} \cdot X_{2,i} + \beta_{prod} \cdot X_{3,i} + \beta_{sympatry} \cdot X_{4,i}$$

The observed co-occurrence $k_i$ is modeled with **Fisher's Hypergeometric distribution** representing local survey plots within that site:

$$k_i \sim \text{FisherHypergeometric}(m_{1i},m_{2i}, N_i, ω_i)$$

where:

- $k_i$ = number of sites where both species co-occur

- $m_{1i}$ = total number of sites where species A is present for pair $i$

- $m_{2i}$ = total number of sites where species B is present for pair $i$

- $N_i$ = total number of sites sampled within the sympatric range of pair $i$

- $ω_i$ = the conditional co-occurrence odds ratio for pair $i$

## Data Generation

I want to simulate data resembling the study design of BBS (\~40km long routes, each having 50 stops). Therefore I need to create two data sets:

1\) route scale (looking at syntopy across transects within the range overlap)

2\) stop scale (looking at syntopy within transects)

For the Fisher's hypergeometric distribution we can use the library `BiasedUrn`, for the modeling we will use `rethinking` and `rstan`.

```{r}
library(BiasedUrn)
library(rethinking)
library(rstan)
set.seed(1)
```

### Route Scale

For the number of sites I choose a \*\*negative binomial \*distribution\*\*, because it is an integer and I know that there is a long tail on the right site of the distribution.

```{r}
#| label: Route Scale Data Simulation
num_pairs <- 100
route_occupancy <- function(num_pairs){
  # simulates the occupancy of species across N routes
  
  # N of routes
  N <-  rnbinom(n = num_pairs, mu = num_pairs, size = 2)
  # Probability distribution for species occupancy
  prob_m1 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  prob_m2 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  # Species occupancy
  m1 <- rbinom(num_pairs, size = N, prob = prob_m1)
  m2 <- rbinom(num_pairs, size = N, prob = prob_m2)
  list(N=N, m1=m1, m2=m2)
}
route_oc <- route_occupancy(num_pairs = num_pairs)

hist(route_oc$N, breaks = 12, main="N Sites Distribution", xlab="Number of sites for sister pair", col="grey")
```

Some species pairs have very large number of sampled sites, while some of them have just a few sampled sites within the sympatric range. We can check the parameters of the distribution:

```{r}
summary(route_oc$N)
```

We can also check the occupancy distribution of the first species set in the pairs:

```{r}
hist(route_oc$m1, breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

This plot nicely resembles species abundance distribution, so if we take a logarithm of it it should be close to normal distribution.

```{r}
hist(log(route_oc$m1), breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

It is exactly like that!

Let's continue with the generation of parameter values. First I will generate three already standardized variables using **normal distribution**.

```{r}
#| label: Parameters
X_res_div   <- rnorm(num_pairs, mean = 0, sd = 1) # X1
X_hab_pref  <- rnorm(num_pairs, mean = 0, sd = 1) # X2
X_prod      <- rnorm(num_pairs, mean = 0, sd = 1) # X3
hist(X_res_div, breaks = 12, main="Resource Use Divergence", xlab="Degree of divergence", col="orange")
```

For the degree of sympatry (% of range overlap) I need to generate specific distribution as x \> 5 (minimum degree for sympatry is typically 5 %) and x \<= 100. As I want the values to have decimals and I need a strict upper limit I have to use **beta distribution**.

```{r}
#| label: Parameter X4
raw_beta <- rbeta(num_pairs, shape1 = 2, shape2 =3)
sym_min <- 5 # sympatry threshold is 5 %
symp_max <- 100 # full overlap
sympatry <- sym_min + (raw_beta - min(raw_beta)) * (symp_max - sym_min) / (max(raw_beta) - min(raw_beta))
hist(sympatry, breaks = 12, main="Degree of Sympatry", xlab="Range overlap (%)", col="darkseagreen3")
```

We can see that most of the pairs have a high degree of sympatry and few pairs have very low values. Now we have to standardize it prior to modeling.

```{r}
#| label: Parameter X4
X_sympatry <- (sympatry - mean(sympatry)) / sd(sympatry) # X3
hist(X_sympatry, breaks = 12, main="zSympatry (Standardized)", xlab="Z-score", col="steelblue")
```

Now we set the "true" values for the parameters in our made-up universe.

```{r}
#| label: True Parameters
beta_0        <- 0.5
beta_res_div  <- 0.8
beta_hab_pref <- -0.6
beta_sympatry <- 1.2
beta_prod     <- 0.3
```

Next step is the calculation of linear combination of parameters and we exponentiate the log-odds to get simple odds ratio.

```{r}
#| label: Odds Ratio
log_omega <- beta_0 + 
             beta_res_div  * X_res_div + 
             beta_hab_pref * X_hab_pref + 
             beta_sympatry * X_sympatry + 
             beta_prod     * X_prod
omega <- exp(log_omega)
```

With this setted up, we calculate the linear predictor for each pair using `BiasedUrn` library.

```{r}
#| label: Simulation of co-occurrence
m1 <- route_oc$m1
m2 <- route_oc$m2
k <- integer(num_pairs)
for(i in 1:num_pairs) {
  # Simple safety boundary: if a species isn't present at all, co-occurrence is 0
  if(m1[i] == 0 || m2[i] == 0) {
    k[i] <- 0
  } else {
    # rFNCHypergeo(n_samples, m1, m2, total_draws, odds)
    k[i] <- rFNCHypergeo(1, route_oc$m1[i], route_oc$N[i] - route_oc$m1[i], route_oc$m2[i], omega[i])
  }
}
```
Let's pack it up into data.frame and take a look!
```{r}
#| label: Route Scale Data Final
route_data <- data.frame(
  N = route_oc$N, m1 = m1, m2 = m2, k = k,
  res_div = X_res_div, hab_pref = X_hab_pref,
  prod = X_prod, 
  sympatry = X_sympatry
)
head(route_data)
```
The data.frame looks good and we should now explore the data before proceeding into next stage.
We need to exclude all pairs with zero co-occurrence before generating point scale data, because for this we need only the routes where both species are present (there is no reason to analyse routes with zero co-occurrence, it will highly bias the results).
```{r}
#| label: Route Scale Data Exploration & Wrangling
nrow(route_data[route_data$k == 0,]) # There is 17 pairs with k=0
route_data[route_data$m1 == 0 & route_data$m2 == 0,] # There is no route with zero occupancy of both species
route_data[abs(route_data$m1 - route_data$m2) >100,] # There is 5 pairs with a very high difference in occupancy between species
```
Higher abundance among sister pairs (e.g. 80 occupied routes from 100, together at 64) can lead to the situation when both species in the pair tend to co-occur more likely simply because they are present everywhere - therefore their co-occurrence affinity will be incomparable with the species which co-occur despite being rare in general (e.g. 5 routes from 100, together at 4). This problem of prevalence can happen if we measure co-occurrence with widely used indicies of similarity (e.g., Jaccard, Simpson).
------------------------------------------------------------------------

### Point Scale

```{r}
#| label: Generate Point-Scale Data Conditioned on Transect Abundance

# 1. Identify which species pairs have at least one co-occurring transect
valid_pairs <- which(route_data$k >= 1)

transect_id       <- c()
parent_pair_id    <- c()
points_total      <- c()
points_m1         <- c()
points_m2         <- c()
points_cooccur_k  <- c()

# 2. Loop through each valid species pair
for (i in valid_pairs) {
  
  num_cooccurring_transects <- route_data$k[i]
  
  # Calculate regional occupancy rates for this specific pair
  regional_rate_A <- route_data$m1[i] / route_data$N[i]
  regional_rate_B <- route_data$m2[i] / route_data$N[i]
  
  # Loop through each of those specific co-occurring transects
  for (t in 1:num_cooccurring_transects) {
    P_total <- 50
    
    # Generate point occupancy probabilities influenced by regional rate
    local_prob_A <- rbeta(1, shape1 = regional_rate_A * 5, shape2 = (1 - regional_rate_A) * 5)
    local_prob_B <- rbeta(1, shape1 = regional_rate_B * 5, shape2 = (1 - regional_rate_B) * 5)
    
    # Boundary protection
    local_prob_A <- max(0.05, min(0.95, local_prob_A))
    local_prob_B <- max(0.05, min(0.95, local_prob_B))
    
    # Draw point-scale occupancy counts bounded by 50 points
    p_m1 <- rbinom(1, size = P_total, prob = local_prob_A)
    p_m2 <- rbinom(1, size = P_total, prob = local_prob_B)
    
    # Safety floor: since they co-occur at the transect scale, they must occupy >= 1 point
    p_m1 <- max(1, p_m1)
    p_m2 <- max(1, p_m2)
    
    # Extract parent pair's regional affinity (omega)
    local_omega <- omega[i] 
    
    # FIXED: Argument names removed to prevent the "formal argument 'n'" collision
    p_k <- rFNCHypergeo(1, p_m1, P_total - p_m1, p_m2, local_omega)
    
    # Append variables
    parent_pair_id   <- c(parent_pair_id, i)
    transect_id      <- c(transect_id, paste0("Pair_", i, "_Transect_", t))
    points_total     <- c(points_total, P_total)
    points_m1        <- c(points_m1, p_m1)
    points_m2        <- c(points_m2, p_m2)
    points_cooccur_k <- c(points_cooccur_k, p_k)
  }
}

# 3. Create long-format point-scale data frame
point_scale_data <- data.frame(
  transect_id = transect_id,
  parent_pair_id = parent_pair_id,
  P_total = points_total,
  m1_points = points_m1,
  m2_points = points_m2,
  k_points = points_cooccur_k
)

head(point_scale_data)
```

# Visualisation

# Modeling

Modeling with Fisher's hypergeometric distribution requires

```{r}
#| include: false
# 1. Prepare your data list (using the variables from your script)
dat_list_fisher <- list(
  k = sim_data$k,
  m1 = sim_data$m1,
  m2 = sim_data$m2,
  N = sim_data$N,
  res_div = sim_data$res_div,
  hab_pref = sim_data$hab_pref,
  sympatry = sim_data$sympatry,
  prod = sim_data$prod
)

# 2. Build the skeleton code using ulam()
# The custom() macro does two things: it writes the "target += ..." Stan call, 
# and it forces ulam to correctly declare k, m1, m2, and N in the data block.
m_skeleton <- ulam(
  alist(
    # Likelihood 
    k ~ custom( fisher_hyper_lpmf(k | exp(log_omega), m1, m2, N) ),
    
    # Linear Predictor
    log_omega <- b0 + b_res*res_div + b_hab*hab_pref + b_sym*sympatry + b_prod*prod,
    
    # Priors
    b0 ~ dnorm(0, 1),
    b_res ~ dnorm(0, 1),
    b_hab ~ dnorm(0, 1),
    b_sym ~ dnorm(0, 1),
    b_prod ~ dnorm(0, 1)
  ), 
  data = dat_list_fisher, 
  sample = FALSE # CRITICAL: This tells ulam to build the code, but not run it!
)

# 3. Extract the autogenerated Stan code
stan_code <- stancode(m_skeleton)

# FIX: Add the missing row-indices [i] for m1, m2, and N that ulam missed
stan_code <- gsub(
  pattern = "fisher_hyper_lpmf(k[i] | exp(log_omega[i]), m1, m2, N)",
  replacement = "fisher_hyper_lpmf(k[i] | exp(log_omega[i]), m1[i], m2[i], N[i])",
  x = stan_code,
  fixed = TRUE
)

# 4. Define your supervisor's C++ function block (with corrected variable names)
functions_block <- "
functions {
  real fisher_hyper_lpmf(int k, real mu, int m1, int m2, int N) {
    if (mu <= 0) return not_a_number();
    
    // Bounds for hypergeometric distribution
    int k_min = max({0, m1 + m2 - N});
    int k_max = min({m1, m2});
    
    // Safety check for support bounds
    if (k < k_min || k > k_max) return negative_infinity();
    
    // Log-numerator for the observed value k
    real log_num = lchoose(m1, k) + lchoose(N - m1, m2 - k) + k * log(mu);
    
    // Normalization constant calculation
    int range_size = k_max - k_min + 1;
    vector[range_size] terms;
    
    for (i in 0:(range_size - 1)) {
      int curr_i = k_min + i;
      terms[i + 1] = lchoose(m1, curr_i) + lchoose(N - m1, m2 - curr_i) + curr_i * log(mu);
    }
    
    return log_num - log_sum_exp(terms);
  }
}
"

# 5. Inject the functions block at the very top of the corrected code
full_stan_code <- paste0(functions_block, stan_code)

# 6. Compile and sample directly with rstan
m_fisher_rethinking <- stan(
  model_code = full_stan_code,
  data = dat_list_fisher,
  chains = 4, 
  cores = 4,
  iter = 2000,
  seed = 12
)
```

```{r}
precis(m_fisher_rethinking)
```

# Smaller number of sister pairs (n=30)

```{r}
sim_data_30 <- sim_data
visualize_simulation(sim_data_30)
```

```{r}
precis(sim_data_30)
```

# Conclusion

# Reference

Remeš, V. and Harmáčková, L. (2023), Resource use divergence facilitates the evolution of secondary syntopy in a continental radiation of songbirds (Meliphagoidea): insights from unbiased co-occurrence analyses. Ecography, 2023: e06268. https://doi.org/10.1111/ecog.06268

Newbury, A. (2024). Bayesian estimation of co-occurrence affinity with dyadic regression. bioRxiv. https://doi.org/10.1101/2024.01.16.575941
