---
title: "Bayesian Analysis of Syntopy"
subtitle: "Assignment for Bayesian Statistics Course"
author: "Matěj Tvarůžka"
date: "`r Sys.Date()`"
categories: [Syntopy, Spatial Data, Birds, R, Bayesian Inference]
format: html
editor: visual
bibliography: references.bib
---

# Introduction

Following the framework of @remes2023, this assignment simulates and analyses local co-occurrence (syntopy) among sister species of birds. In evolutionary ecology, syntopy is the final stage of the speciation cycle, during which two sister species meet at the same sites within their range overlap. I model the co-occurrence probability at local sites as a function of ecological and spatial predictors. This idea is connected to the topic of my master’s thesis, in which I model syntopy at two spatial scales (routes and stops) using BBS (Breeding Bird Survey) data for passerines and eBird Status & Trends range maps.

::: callout-note
## Assignment Goals

1.  Generate dream syntopy data for both scales (routes and stops).
2.  Visualise and explore generated data.
3.  Fit models for both scales using the `ulam` function and compare the estimates with true values.
:::

To measure syntopy accurately, we need to consider only sites within the range overlap for the analysis, because otherwise it will lead to biased results @remes2023. We can imagine the range overlap being the jar, from which we randomly draw balls (sites). There are four possible outcomes of that draw:

1.  Both species are present,
2.  Species A is present,
3.  Species B is present,
4.  Both species are missing at the site.

If we want to model this problem using Bayesian methods, we can follow the approach in @newbury measuring the degree of species co-occurrence with the log-odds ratio (see box 1). The most accurate probability distribution for our problem is Fisher’s non-central hypergeometric distribution, because the sampling is carried out without replacement (unlike the binomial distribution), which reflects the true nature of the problem, given the limited number of pairs.

::: callout-note
## Box 1: Bayesian Log Odds Ratio

The log-odds ratio serves as the foundational metric to quantify spatial association between sister species. Based on a standard $2 \times 2$ presence-absence matrix ($a$ = both present, $b/c$ = single species present, $d$ = both absent), the empirical calculation is:

$$\log \text{OR}_{\text{}} = \ln\left(\frac{a \times d}{b \times c}\right)$$

In this assignment, we transition away from this traditional empirical formula and instead implement a Bayesian framework. The key distinctions and advantages are:

1.  **Mathematical Formulation:**

    The observed co-occurrence $k_i$ is modelled with Fisher's Hypergeometric distribution representing local survey plots within that site: $$k_i \sim \text{FisherHypergeometric}(m_{1i}, m_{2i}, N_i, \alpha_i)$$

    where:

    - $k_i$ = number of sites where both species co-occur

    - $m_{1i}$ = total number of sites where species A is present for pair $i$

    - $m_{2i}$ = total number of sites where species B is present for pair $i$

    - $N_i$ = total number of sites sampled within the sympatric range of pair $i$

    - $\alpha_i$ = the conditional co-occurrence odds ratio for pair $i$.

    For each site $i$ out of $N$, the co-occurrence affinity ($ω_i$) is the log-odds ratio of given parameters:

    $$\text{log}(\alpha_i) = \beta_{0} + \beta_{1} \cdot X_{1,i} + \lambda_{1,i}$$

2.  **Mathematical Stability:** By evaluating data through the Log Probability Mass Function (LPMF) of Fisher's Non-Central Hypergeometric distribution, our model naturally handles small sample sizes and zero counts. It completely bypasses the need to delete data or inject mathematical bandages.

3.  **Symmetric Linear Predictors:** Taking the natural logarithm maps highly asymmetric raw odds ratios onto a continuous linear space from $-\infty$ to $+\infty$, centering independent sorting perfectly at $0$ ($>0$ is syntopy, $<0$ is segregation). This allows our ecological parameters ($\beta_{\text{res}}$, $\beta_{\text{hab}}$, etc.) to be added together seamlessly.

4.  **Information Pooling:** Unlike the typical formula, which treats every route as an isolated island, our hierarchical structure links local routes together. If a specific route has noisy counts, the model performs **partial pooling**—shrinking its local estimate back toward the regional baseline dictated by the environmental covariates.

5.  **Prevalence Invariance:** Unlike traditional similarity indices (e.g., Jaccard, Simpson), this metric standardises for baseline regional abundance, therefore isolates true ecological affinity without being biased by widespread or ubiquitous species.
:::

# Data Simulation

To satisfy the assignment's complexity requirements, I define **5 true parameters** that dictate the data-generating process:

- $\beta_{0} = 0.5$ (Baseline log-odds of co-occurrence; slight facilitation)

- $\beta_{res.div} = -0.7$ (Resource use divergence: greater divergence increases syntopy)

- $\beta_{hab.pref} = 0.6$ (Difference in micro-habitat and environmental preference decreases syntopy)

- $\beta_{prod} = 0.3$ (Environmental productivity (e.g., NDVI): richer environments support co-occurrence)

- $\beta_{sympatry} = 0.8$ (Spatial overlap: regional overlap strongly increases syntopy)

I want to simulate data that resembles the BBS study design (\~40 km-long routes, each with 50 stops). Therefore, I need to create two data sets:

1\) route scale (looking at syntopy across transects within the range overlap)

2\) stop scale (looking at syntopy within transects)

For Fisher's non-central hypergeometric distribution we can use the library `BiasedUrn` library, and for the modelling, we will use libraries `rethinking` and `rstan`.

```{r}
#| include: false
library(BiasedUrn)
library(rethinking)
library(rstan)
library(tidyverse)
set.seed(1)
```

## Route Scale

For the number of sites, I choose a negative binomial distribution because it is integer-valued and has a long right tail.

```{r}
#| label: Route Scale Data Simulation
num_pairs <- 100
route_occupancy <- function(num_pairs){
  # simulates the occupancy of species across N routes
  
  # N of routes
  N <-  rnbinom(n = num_pairs, mu = num_pairs, size = 2)
  # Probability distribution for species occupancy
  prob_m1 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  prob_m2 <- rbeta(num_pairs, shape1 = 1, shape2 = 2)
  # Species occupancy
  m1 <- rbinom(num_pairs, size = N, prob = prob_m1)
  m2 <- rbinom(num_pairs, size = N, prob = prob_m2)
  list(N=N, m1=m1, m2=m2)
}
route_oc <- route_occupancy(num_pairs = num_pairs)

hist(route_oc$N, breaks = 12, main="N Sites Distribution", xlab="Number of sites for sister pair", col="purple")
```

Some species pairs have a very large number of sampled sites, while others have just a few within the sympatric range. We can check the parameters of the distribution:

```{r}
summary(route_oc$N)
```

We can also check the occupancy distribution of the first species set in the pairs:

```{r}
hist(route_oc$m1, breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

This plot resembles a species abundance distribution, so if we take the logarithm of it, it should be close to a normal distribution.

```{r}
hist(log(route_oc$m1), breaks = 12, main="Species Occupancy Sample", xlab="Number of occupied routes", col="red")
```

It is exactly like that!

Let us continue with parameter generation. First, I will generate three already standardised variables using the normal distribution.

```{r}
X_res_div   <- rnorm(num_pairs, mean = 0, sd = 1) # X1
X_hab_pref  <- rnorm(num_pairs, mean = 0, sd = 1) # X2
X_prod      <- rnorm(num_pairs, mean = 0, sd = 1) # X3
hist(X_res_div, breaks = 12, main="Resource Use Divergence", xlab="Degree of divergence", col="orange")
```

Let us continue with parameter generation. First, I will generate three already standardised variables using the normal distribution. For the degree of sympatry (% of range overlap), I need to generate a distribution with x \> 5 (the minimum degree of sympatry is typically 5 %) and x \<= 100. As I want the values to have decimals and I need a strict upper limit, I have to use a beta distribution.

```{r}
raw_beta <- rbeta(num_pairs, shape1 = 2, shape2 =3)
sym_min <- 5 # sympatry threshold is 5 %
symp_max <- 100 # full overlap
sympatry <- sym_min + (raw_beta - min(raw_beta)) * (symp_max - sym_min) / (max(raw_beta) - min(raw_beta))
hist(sympatry, breaks = 12, main="Degree of Sympatry", xlab="Range overlap (%)", col="darkseagreen3")
```

We can see that most pairs have a high degree of sympatry, while a few have very low values. Now we have to standardise it prior to modelling.

```{r}
#| label: Parameter X4
X_sympatry <- (sympatry - mean(sympatry)) / sd(sympatry) # X3
hist(X_sympatry, breaks = 12, main="zSympatry (Standardised)", xlab="Z-score", col="steelblue")
```

Now we set the "true" parameter values in our made-up universe.

```{r}
#| label: True Parameters
beta_0        <- 0.5
beta_res_div  <- -0.7
beta_hab_pref <- 0.6
beta_sympatry <- 0.8
beta_prod     <- 0.3
```

The next step is to calculate a linear combination of parameters, and we exponentiate the log-odds to obtain a simple odds ratio.

```{r}
#| label: Linear Combination of True Parameters
log_alpha <- beta_0 + 
             beta_res_div  * X_res_div + 
             beta_hab_pref * X_hab_pref + 
             beta_sympatry * X_sympatry + 
             beta_prod     * X_prod
alpha <- exp(log_alpha) # Odds Ratio
```

With this settled, we calculate the linear predictor for each pair using the `BiasedUrn` library.

```{r}
#| label: Simulation of Co-occurrence
m1 <- route_oc$m1
m2 <- route_oc$m2
k <- integer(num_pairs)
for(i in 1:num_pairs) {
  set.seed(1)
  # Simple safety boundary: if a species isn't present at all, co-occurrence is 0
  if(m1[i] == 0 || m2[i] == 0) {
    k[i] <- 0
  } else {
    # rFNCHypergeo(n_samples, m1, m2, total_draws, odds)
    k[i] <- rFNCHypergeo(1, route_oc$m1[i], route_oc$N[i] - route_oc$m1[i], route_oc$m2[i], alpha[i])
  }
}
```

Let us pack it up into `data.frame` and take a look at the data.

```{r}
#| label: Route Scale Data Final
route_data <- data.frame(
  N = route_oc$N, m1 = m1, m2 = m2, k = k,
  res_div = X_res_div, hab_pref = X_hab_pref,
  prod = X_prod, 
  sympatry = X_sympatry
)
head(route_data)
```

The `data.frame` looks good, and we should now explore the data before proceeding to the next stage.

```{r}
#| label: Route Scale Data Exploration & Wrangling
nrow(route_data[route_data$k == 0,])
# There is 18 pairs with k=0
route_data[abs(route_data$m1 - route_data$m2) >100,]
# There are 4 pairs with a very high difference in occupancy between species
```

Now we have to make an unique id for each pair to be able to link it later with stops data.

```{r}
# Add unique IDs, true odds ratio and true log-odds ratio
route_data$pair_id <- 1:nrow(route_data)
route_data$OR <- alpha
route_data$log_OR <- log_alpha
route_data[route_data$m1 == 0 & route_data$m2 == 0,]
```

There are 2 routes with zero occupancy, so-called double zeros. Double zeros could mean two things. Either both species are truly absent at the route (true negatives), or they were not detected during the survey (false negatives). Given that the route is within their range zone and there were 50 points where at least one of the species could have been detected, it should be more likely that they were not there.

## Stop Scale

The next step is to simulate point-scale data from the route data we already have. I choose to filter only pairs with **k \>= 1**, because there is no reason to calculate syntopy on transects where there is 0 possibility of co-occurrence.

```{r}
# Select only rows where k >= 1
valid_route_data <- route_data[route_data$k >= 1, ]

# Initialize empty list and counter for fast execution
datalist <- list()
counter <- 1

for (row_idx in 1:nrow(valid_route_data)) {
  
  # Extract pair metadata and parameters
  current_pair  <- valid_route_data$pair_id[row_idx]
  num_routes <- valid_route_data$k[row_idx]
  local_alpha   <- valid_route_data$OR[row_idx] # using same alpha as for the route sclae
  
  # Calculate the realized regional occupancy rates for this specific pair
  rate_A <- valid_route_data$m1[row_idx] / valid_route_data$N[row_idx]
  rate_B <- valid_route_data$m2[row_idx] / valid_route_data$N[row_idx]
  
  # Loop through each of those specific co-occurring transects
  for (t in 1:num_routes) {
    S_total <- 50
    
    # Generate local point occupancy probabilities using the pair's regional rates
    local_prob_A <- rbeta(1, shape1 = rate_A * 5, shape2 = (1 - rate_A) * 5)
    local_prob_B <- rbeta(1, shape1 = rate_B * 5, shape2 = (1 - rate_B) * 5)
    
    # Boundary protection to prevent 0 or 1 probabilities
    local_prob_A <- max(0.01, min(0.99, local_prob_A))
    local_prob_B <- max(0.01, min(0.99, local_prob_B))
    
    # Draw stop-scale occupancy counts bounded by 50 points
    s_m1 <- rbinom(1, size = S_total, prob = local_prob_A)
    s_m2 <- rbinom(1, size = S_total, prob = local_prob_B)
    
    # Safety floor: since they co-occurred on this transect, they must occupy >= 1 point
    s_m1 <- max(1, s_m1)
    s_m2 <- max(1, s_m2)
    
    # Draw realized stop co-occurrences using Fisher's Noncentral Hypergeometric
    s_k <- rFNCHypergeo(1, s_m1, S_total - s_m1, s_m2, local_alpha)
    
    # Wrap it up in the list
    datalist[[counter]] <- data.frame(
      route_id = paste0("Pair_",current_pair, "_Route_", t),
      pair_id = current_pair,
      S_total = S_total,
      m1_stops = s_m1,
      m2_stops = s_m2,
      k_stops = s_k
    )
    counter <- counter + 1
  }
}

data <- do.call(rbind, datalist)
head(data)
```

Finally, we join our `data` with the generated parameters for each pair we already have in `route_data`.

```{r}
# Start with the simulated data (which naturally excludes k = 0 pairs)
stop_data <- data |>
  select(pair_id, route_id, S_total, m1_stops, m2_stops, k_stops) |> 
  left_join(
    # Pull ONLY the environmental covariates from route_data
    route_data |> select(pair_id, res_div, hab_pref, prod, sympatry), 
    by = "pair_id"
  )

# Verify the clean data frame
head(stop_data)
```

# Models

We have prepared our data and can now run some models using the `ulam()` function from the `rethinking` package. There are several steps we need to take to get the output:

1.  Wrap up the data into a list so that `ulam` can read it.

2.  Import the separate file for Fisher's non-central hypergeometric likelihood function.

3.  Set up the model formula in `ulam`.

4.  Transform the code into stan syntax.

5.  Run the model using `stan()` function from `rstan` package.

::: {.callout-tip collapse="false"}
## Box 2

The mathematical engine of this analysis is the custom **Fisher's Non-Central Hypergeometric Log Probability Mass Function (`fisher_hyper_lpmf`)**. Unlike a standard hypergeometric distribution, which assumes species are sorted completely at random, this distribution accounts for ecological attraction or repulsion via an odds-ratio parameter.

The function first calculates the absolute physical boundaries of co-occurrence ($k$) allowed by the individual species' occupancy counts ($m_1, m_2$) within the total sampled sites ($N$): $$k_{\min} = \max(0, m_1 + m_2 - N)$$ $$k_{\max} = \min(m_1, m_2)$$

If an observed or sampled co-occurrence count falls outside this range ($k < k_{\min}$ or $k > k_{\max}$), the function returns $-\infty$, assigning it a probability of 0 and rendering that parameter space impossible for the MCMC sampler.
:::

## Route Scale Model

For each route $i$ out of $N$, the co-occurrence affinity ($α$) is the log-odds ratio of given parameters:

$$\text{log}(\alpha_i) = \beta_{0} + \beta_{res.div} \cdot X_{1,i} + \beta_{hab.pref} \cdot X_{2,i} + \beta_{prod} \cdot X_{3,i} + \beta_{sympatry} \cdot X_{4,i}$$

```{r}
#| echo: true
#| message: false
#| warning: false
#| paged-print: false
dat_list <- list(
  k = route_data$k,
  m1 = route_data$m1,
  m2 = route_data$m2,
  N = route_data$N,
  res_div = route_data$res_div,
  hab_pref = route_data$hab_pref,
  prod = route_data$prod,
  sympatry = route_data$sympatry
)
# Read separate stan file and wrap it in the required Stan syntax
likelihood_pmf <- paste(
  "functions {",
  paste(readLines("fisher_nchyper_lpmf.stan"), collapse = "\n"),
  "}",
  sep = "\n"
)

# Model settings
model <- ulam(
  alist(
    # Likelihood
    k ~ custom( fisher_hyper_lpmf(k[i] | exp(log_alpha[i]), m1[i], m2[i], N[i]) ),
    
    # Linear Predictor
    log_alpha <- b0 + b_res*res_div + b_hab*hab_pref + b_prod*prod + b_sym*sympatry,
    
    # Priors
    b0 ~ dnorm(0, 3),
    b_res ~ dnorm(0, 3),
    b_hab ~ dnorm(0, 3),
    b_sym ~ dnorm(0, 3),
    b_prod ~ dnorm(0, 3)
  ), 
  data = dat_list, 
  sample = FALSE # Generate text for stan, not sampling yet
)

# Extract the autogenerated text
stan_code <- stancode(model)

# Paste the file-loaded functions block to the top of the text string
full_stan_code <- paste0(likelihood_pmf, "\n", stan_code)

# Run the compiled model using rstan
out_routes <- stan(
  model_code = full_stan_code,
  data = dat_list,
  chains = 4, 
  cores = 4,
  iter = 4000,
  seed = 1,
  refresh = 0 # <--- Turns off the iteration progress tracker
)
```

First, we can check the convergence of chains.

```{r}
traceplot(out_routes, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod"))
```

All chains have converged nicely!

```{r}
precis(out_routes)
```

Our model estimates for most parameters are very good; the intercept mean (baseline log-odds ratio) is slightly underestimated, despite the credible interval including the true value (0.5).

## Stop Scale Model

```{r}
#| message: false
#| warning: false
# Map character route IDs to sequential integers for Stan indexing
stop_data$route_index <- as.integer(as.factor(stop_data$route_id))

dat_list_stops <- list(
  k_stops     = stop_data$k_stops,
  m1_stops    = stop_data$m1_stops,
  m2_stops    = stop_data$m2_stops,
  S_total     = stop_data$S_total,
  res_div     = stop_data$res_div,
  hab_pref    = stop_data$hab_pref,
  prod        = stop_data$prod,
  sympatry    = stop_data$sympatry,
  route_index = stop_data$route_index
)

# Model Settings
model_stops <- ulam(
  alist(
    # Likelihood using stop-level data arrays
    k_stops ~ custom( fisher_hyper_lpmf(k_stops[i] | exp(log_omega[i]), m1_stops[i], m2_stops[i], S_total[i]) ),
    
    # Linear Predictor: Pair baseline + Route random effect deviation
log_omega <- b0 + b_res*res_div + b_hab*hab_pref + b_prod*prod + b_sym*sympatry + (z_route[route_index] * sigma_route),

    # Fixed Effect Priors
   b0 ~ dnorm(0, 3),
    b_res ~ dnorm(0, 3),
    b_hab ~ dnorm(0, 3),
    b_sym ~ dnorm(0, 3),
    b_prod ~ dnorm(0, 3),
    
    # Adaptive Priors
    # Non-Centered Random Effect Layer
    # z_route is a completely standard, clean vector array
    z_route[route_index] ~ dnorm(0, 1),
    sigma_route ~ dexp(1)
    # exponential distribution forces sigma_route
    # to be strictly positive (since a standard deviation cannot be negative)
  ), 
  data = dat_list_stops, 
  sample = FALSE # Generate text for stan, not sampling yet
)

# Extract the autogenerated text
stan_code_stops <- stancode(model_stops)

# Paste the file-loaded functions block to the top of the text string
full_stan_code_stops <- paste0(likelihood_pmf, "\n", stan_code_stops)

# Run the compiled model using rstan
out_stops <- stan(
  model_code = full_stan_code_stops,
  data = dat_list_stops,
  chains = 4, 
  cores = 4,
  iter = 4000,
  seed = 1,
  refresh = 0 # <--- Turns off the iteration progress tracker
)
```

There were several warnings at the and of the sampling, that some chains did not converged in an optimal way so we first check the traceplot for the main parameters.

```{r}
traceplot(out_stops, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod", "sigma_route"))
```

Now we can take a look at the parameter estimates.

```{r}
precis(out_stops, pars = c("b0", "b_res", "b_hab", "b_sym", "b_prod", "sigma_route"))
```

```{r}
# Extract posterior sample draws
post_stops <- extract.samples(out_stops)
n_samples  <- length(post_stops$b0)
n_rows     <- nrow(stop_data)

# Create an empty matrix to hold log-odds (rows = post-samples, cols = stop_data rows)
log_omega_routes <- matrix(NA, nrow = n_samples, ncol = n_rows)

# Reconstruct log_omega per transect by pulling pair fixed effects and adding local random noise
for(i in 1:n_rows) {
  pair_base <- post_stops$b0 + 
               post_stops$b_res * stop_data$res_div[i] + 
               post_stops$b_hab * stop_data$hab_pref[i] + 
               post_stops$b_sym * stop_data$sympatry[i] + 
               post_stops$b_prod * stop_data$prod[i]
  
  # Manually reconstruct the route deviation from the non-centered samples
  current_route_idx <- stop_data$route_index[i]
  reconstructed_a_route <- post_stops$z_route[, current_route_idx] * post_stops$sigma_route
  
  # Total local log-odds ratio
  log_omega_routes[, i] <- pair_base + reconstructed_a_route
}

# 1. Compute and assign the estimated log-odds ratio back to each individual transect row
stop_data$estimated_local_log_OR <- colMeans(log_omega_routes)

# 2. Average the estimated log-odds values for each distinct pair_id
pair_level_averages <- aggregate(estimated_local_log_OR ~ pair_id, data = stop_data, FUN = mean)
print(head(pair_level_averages))
```

We can compare it with the true values of local logOR.

```{r}
stop_model_evaluation <- pair_level_averages %>% 
  left_join(route_data %>% select(pair_id, true_log_OR = log_OR), by = "pair_id")
print(head(stop_model_evaluation))
```

# Conclusion

# Reference
