---
title: "Bayesian Analysis of Syntopy"
author: "Matěj Tvarůžka"
date: "`r Sys.Date()`"
format: html
editor: visual
---

# Introduction

Following the framework of Remeš & Harmáčková (2023), this assignment simulates and analyses local co-occurrence (syntopy) among sister species of birds. In evolutionary ecology, syntopy is the final stage of the speciation cycle, during which two sister species meet at the same sites. I model the co-occurrence probability at local sites as a function of ecological and spatial predictors. This idea is connected to the topic of my master's thesis, where I model syntopy at two spatial scales using BBS data for passerines and range maps from eBird status & trends.

To measure syntopy accurately, we need to consider only sites within the range overlap for the analysis. We can imagine the range overlap being the jar, from which we randomly draw sampled sites. There are four possible outcomes of that draw: 1) both species are present, 2) species A is present, 3) species B is present, or 4) both species are missing at the site. To measure the degree of species co-occurrence, we use the log-odds ratio because it centres the odds ratio around zero, making the scale linear.

If we want to model this problem using Bayesian methods, we can follow the approach in Newbury (2024), but we will not include random effects in the model and will use only a simple Bayesian GLM. The most accurate probability distribution for our problem is **Fisher's non-central hypergeometric distribution**, where *N* denotes the population size (number of sites within the sympatric zone), *K* is the number of co-occurrence states (*K* = 4), *n* is the number of draws, and *k* is the number of co-occurrences. This distribution is accurate because the sampling is carried out without replacement (unlike the binomial distribution), which reflects the true nature of the problem, given the limited number of pairs.

::: callout-note
## Goals

1.  Generate dream data for 100 pairs using Fisher's hypergeometric distribution.
2.  Visualize the generated data.
3.  Fit the model using `ulam` function and compare the estimates using different N.
:::

# Generative Process

To satisfy the assignment's complexity requirements, I define **5 true parameters** that dictate the data-generating process:

- $\beta_{0} = 0.5$ (Baseline log-odds of co-occurrence; slight facilitation)

- $\beta_{res.div} = -0.8$ (Resource use divergence: greater divergence decreases syntopy)

- $\beta_{hab.pref} = 0.6$ (Similar microhabitat and environmental preference increases syntopy)

- $\beta_{sympatry} = 1.2$ (Spatial overlap: regional overlap strongly increases syntopy)

- $\beta_{prod} = 0.3$ (Environmental productivity (e.g., NDVI): richer environments support co-occurrence)

## Fishers' Hypergeometric Distribution

### Mathematical Formulation

For each site $i$ out of $N$, the co-occurrence affinity ($α$) is the log-odds ratio of given parameters:

$$\text{log}(ω_i) = \beta_{0} + \beta_{res.div} \cdot X_{1,i} + \beta_{hab.pref} \cdot X_{2,i} + \beta_{sympatry} \cdot X_{3,i} + \beta_{prod} \cdot X_{4,i}$$

The observed co-occurrence $k_i$ is modeled with **Fisher's Hypergeometric distribution** representing local survey plots within that site:

$$k_i \sim \text{FisherHypergeometric}(m_{1i},m_{2i}, N_i, ω_i)$$

where:

- $k_i$ = number of sites where both species co-occur

- $m_{1i}$ = total number of sites where species A is present for pair $i$

- $m_{2i}$ = total number of sites where species B is present for pair $i$

- $N_i$ = total number of sites sampled within the sympatric range of pair $i$

- $ω_i$ = the conditional co-occurrence odds ratio for pair $i$

### Data Generation

For the Fisher's hypergeometric distribution we can use the library `BiasedUrn`

```{r}
#| include: false
library(BiasedUrn)
library(rethinking)
```

For the number of sites I choose a \*\*gamma-Poisson \*distribution\*\*, because it is an integer and I know that there is a long tail on the right site of the distribution.

```{r}
#| label: Data generation for 100 pairs FHD
set.seed(12)
num_pairs <- 30
N <- rpois(num_pairs, lambda = rgamma(n = num_pairs, shape = 2, rate = 1/100)) # gamma-Poisson distribution, N of sites
m1 <- rbinom(num_pairs, size = N, prob = 0.5) # sites with species A
m2 <- rbinom(num_pairs, size = N, prob = 0.5) # sites with species B
hist(N, breaks = 12, main="N Sites Distribution", xlab="Number of sites for sister pair", col="red")
```

Some species pairs have very large number of sampled sites, while some of them have just a few sampled sites within the sympatric range.

```{r}
summary(N)
```

Let's continue with the generation of parameter values. First I will generate three already standardized variables using **normal distribution**.

```{r}
set.seed(12)
X_res_div   <- rnorm(num_pairs, mean = 0, sd = 1) # X1
X_hab_pref  <- rnorm(num_pairs, mean = 0, sd = 1) # X2
X_prod      <- rnorm(num_pairs, mean = 0, sd = 1) # X4
hist(X_res_div, breaks = 12, main="Resource Use Divergence", xlab="Degree of divergence", col="orange")
```

For the parameter X3 (% of range overlap) I need to generate specific distribution as x \> 5 (minimum degree for sympatry is typically 5 %) and x \<= 100. As I want the values to have decimals and I need a strict upper limit I have to use **beta distribution**.

```{r}
set.seed(12)
raw_beta <- rbeta(num_pairs, shape1 = 2, shape2 = 1)
target_min <- 5 # sympatry treshold is 5 %
target_max <- 100 # full overlap
sympatry <- target_min + (raw_beta - min(raw_beta)) * (target_max - target_min) / (max(raw_beta) - min(raw_beta))
hist(sympatry, breaks = 12, main="Degree of Sympatry", xlab="Range overlap (%)", col="darkseagreen3")
```

We can see that most of the pair have a high degree of sympatry and few pair have very low values. Now we have to standardize it prior to modeling.

```{r}
X_sympatry <- (sympatry - mean(sympatry)) / sd(sympatry) # X3
hist(X_sympatry, breaks = 12, main="zSympatry (Standardized)", xlab="Z-score", col="steelblue")
```

Now we set the "true" values for the parameters in our ideal universe.

```{r}
beta_0        <- 0.5
beta_res_div  <- -0.8
beta_hab_pref <- 0.6
beta_sympatry <- 1.2
beta_prod     <- 0.3
```

Next step is the calculation of linear combination of parameters and we exponentiate the log-odds to get simple odds ratio.

```{r}
log_omega <- beta_0 + 
             beta_res_div  * X_res_div + 
             beta_hab_pref * X_hab_pref + 
             beta_sympatry * X_sympatry + 
             beta_prod     * X_prod
omega <- exp(log_omega)
```

With this setted up, we calculate the linear predictor for each pair using `BiasedUrn` library.

```{r}
set.seed(12)
k <- integer(num_pairs)
for(i in 1:num_pairs) {
  # Simple safety boundary: if a species isn't present at all, co-occurrence is 0
  if(m1[i] == 0 || m2[i] == 0) {
    k[i] <- 0
  } else {
    # rFNCHypergeo(n_samples, m1, m2, total_draws, odds)
    # Note: BiasedUrn expects: m1 = white balls, m2 = black balls, n = balls drawn
    k[i] <- rFNCHypergeo(1, m1[i], N[i] - m1[i], m2[i], omega[i])
  }
}
```

```{r}
sim_data <- data.frame(
  N = N, m1 = m1, m2 = m2, k = k,
  res_div = X_res_div, hab_pref = X_hab_pref, 
  sympatry = X_sympatry, prod = X_prod
)

head(sim_data)
```

------------------------------------------------------------------------

# Data Visualization
```{r}
visualize_simulation <- function(df) {
  # Get the number of pairs dynamically from the data frame
  n_pairs <- nrow(df)
  
  # --- STEP 1: Calculate the Empirical Log-Odds Ratio (Syntopy Outcome) ---
  # Using the Haldane-Anscombe correction (+0.5) to handle zeros safely
  a <- df$k
  b <- df$m1 - df$k
  c <- df$m2 - df$k
  d <- df$N - df$m1 - df$m2 + df$k
  emp_log_OR <- log(((a + 0.5) * (d + 0.5)) / ((b + 0.5) * (c + 0.5)))
  
  # --- STEP 2: Dynamically Adjust Plotting Parameters Based on Sample Size ---
  if (n_pairs < 50) {
    # Small dataset settings
    pt_col   <- rgb(0.2, 0.5, 0.7, 0.9)  # Solid colors, high visibility
    pt_cex   <- 1.3                       # Larger points
    pt_pch   <- 21                        # Open circles with borders
    bg_col   <- "lightblue"               # Fill color for pch 21
    h_breaks <- 8                         # Fewer histogram bins
  } else if (n_pairs >= 50 & n_pairs < 300) {
    # Medium dataset settings (Default)
    pt_col   <- rgb(0.2, 0.5, 0.7, 0.6)  # Moderate transparency
    pt_cex   <- 1.0                       # Standard size
    pt_pch   <- 16                        # Solid circles
    bg_col   <- NA
    h_breaks <- 15                        # Standard bins
  } else {
    # Large dataset settings (300+ pairs) to combat overplotting
    pt_col   <- rgb(0.1, 0.4, 0.6, 0.2)  # High transparency (shows density clusters)
    pt_cex   <- 0.6                       # Smaller points
    pt_pch   <- 16                        # Solid circles
    bg_col   <- NA
    h_breaks <- 30                        # Fine-grained histogram bins
  }
  
  # --- STEP 3: Generate the Multi-Panel Layout ---
  # Set up a 2x2 plotting grid
  par(mfrow = c(2, 2), mar = c(4.5, 4.5, 2.5, 1.5), las = 1)
  
  # Plot 1: Histogram of Sample Sizes (N)
  hist(df$N, breaks = h_breaks, col = "grey85", border = "grey40",
       main = paste("Distribution of N (", n_pairs, " Pairs )", sep=""),
       xlab = "Number of Sites available (N)", ylab = "Frequency")
  
  # Plot 2: Single-Species Occupancy vs Realized Overlap
  plot(df$m1 / df$N, df$k / df$N, col = pt_col, cex = pt_cex, pch = pt_pch, bg = bg_col,
       xlim = c(0, 1), ylim = c(0, 1),
       main = "Realized vs. Marginal Occupancy",
       xlab = "Occupancy Rate of Species A (m1 / N)",
       ylab = "Co-occurrence Rate (k / N)")
  abline(a = 0, b = 1, lty = 2, col = "red") # Theoretical upper limit limit
  
  # Plot 3: Syntopy Outcome vs. Sympatry (Positive Predictor)
  plot(df$sympatry, emp_log_OR, col = pt_col, cex = pt_cex, pch = pt_pch, bg = bg_col,
       main = "Outcome vs. Sympatry Gradient",
       xlab = "Standardized Sympatry (Z-score)",
       ylab = "Empirical Log-Odds Ratio")
  abline(h = 0, lty = 3, col = "grey50")
  # Add trendline
  fit_sym <- lm(emp_log_OR ~ df$sympatry)
  abline(fit_sym, col = "darkorange2", lwd = 2)
  
  # Plot 4: Syntopy Outcome vs. Resource Divergence (Negative Predictor)
  plot(df$res_div, emp_log_OR, col = pt_col, cex = pt_cex, pch = pt_pch, bg = bg_col,
       main = "Outcome vs. Resource Divergence",
       xlab = "Standardized Resource Divergence (Z-score)",
       ylab = "Empirical Log-Odds Ratio")
  abline(h = 0, lty = 3, col = "grey50")
  # Add trendline
  fit_res <- lm(emp_log_OR ~ df$res_div)
  abline(fit_res, col = "firebrick3", lwd = 2)
  
  # Reset plotting parameters to default
  par(mfrow = c(1, 1))
}
visualize_simulation(sim_data)
```


# Modeling
Modeling with Fisher's hypergeometric distribution requires
```{r}
#| include: false
# 1. Prepare your data list (using the variables from your script)
dat_list_fisher <- list(
  k = sim_data$k,
  m1 = sim_data$m1,
  m2 = sim_data$m2,
  N = sim_data$N,
  res_div = sim_data$res_div,
  hab_pref = sim_data$hab_pref,
  sympatry = sim_data$sympatry,
  prod = sim_data$prod
)

# 2. Build the skeleton code using ulam()
# The custom() macro does two things: it writes the "target += ..." Stan call, 
# and it forces ulam to correctly declare k, m1, m2, and N in the data block.
m_skeleton <- ulam(
  alist(
    # Likelihood 
    k ~ custom( fisher_hyper_lpmf(k | exp(log_omega), m1, m2, N) ),
    
    # Linear Predictor
    log_omega <- b0 + b_res*res_div + b_hab*hab_pref + b_sym*sympatry + b_prod*prod,
    
    # Priors
    b0 ~ dnorm(0, 1),
    b_res ~ dnorm(0, 1),
    b_hab ~ dnorm(0, 1),
    b_sym ~ dnorm(0, 1),
    b_prod ~ dnorm(0, 1)
  ), 
  data = dat_list_fisher, 
  sample = FALSE # CRITICAL: This tells ulam to build the code, but not run it!
)

# 3. Extract the autogenerated Stan code
stan_code <- stancode(m_skeleton)

# FIX: Add the missing row-indices [i] for m1, m2, and N that ulam missed
stan_code <- gsub(
  pattern = "fisher_hyper_lpmf(k[i] | exp(log_omega[i]), m1, m2, N)",
  replacement = "fisher_hyper_lpmf(k[i] | exp(log_omega[i]), m1[i], m2[i], N[i])",
  x = stan_code,
  fixed = TRUE
)

# 4. Define your supervisor's C++ function block (with corrected variable names)
functions_block <- "
functions {
  real fisher_hyper_lpmf(int k, real mu, int m1, int m2, int N) {
    if (mu <= 0) return not_a_number();
    
    // Bounds for hypergeometric distribution
    int k_min = max({0, m1 + m2 - N});
    int k_max = min({m1, m2});
    
    // Safety check for support bounds
    if (k < k_min || k > k_max) return negative_infinity();
    
    // Log-numerator for the observed value k
    real log_num = lchoose(m1, k) + lchoose(N - m1, m2 - k) + k * log(mu);
    
    // Normalization constant calculation
    int range_size = k_max - k_min + 1;
    vector[range_size] terms;
    
    for (i in 0:(range_size - 1)) {
      int curr_i = k_min + i;
      terms[i + 1] = lchoose(m1, curr_i) + lchoose(N - m1, m2 - curr_i) + curr_i * log(mu);
    }
    
    return log_num - log_sum_exp(terms);
  }
}
"

# 5. Inject the functions block at the very top of the corrected code
full_stan_code <- paste0(functions_block, stan_code)

# 6. Compile and sample directly with rstan
m_fisher_rethinking <- stan(
  model_code = full_stan_code,
  data = dat_list_fisher,
  chains = 4, 
  cores = 4,
  iter = 2000,
  seed = 12
)
```

```{r}
precis(m_fisher_rethinking)
```

# Smaller number of sister pairs (n=30)

```{r}
sim_data_30 <- sim_data
visualize_simulation(sim_data_30)
```

```{r}
precis(sim_data_30)
```
# Conclusion

# Reference

Remeš, V. and Harmáčková, L. (2023), Resource use divergence facilitates the evolution of secondary syntopy in a continental radiation of songbirds (Meliphagoidea): insights from unbiased co-occurrence analyses. Ecography, 2023: e06268. https://doi.org/10.1111/ecog.06268

Newbury, A. (2024). Bayesian estimation of co-occurrence affinity with dyadic regression. bioRxiv. https://doi.org/10.1101/2024.01.16.575941
